---
title: "The Score is still Q to 12"
subtitle: "Analyzing the Most Chaotic Sport in the Universe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Score is still Q to 12}

  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

# Comic book color palette
comic_colors <- c("#FF6B6B", "#4ECDC4", "#FFD93D", "#6C5CE7", 
                  "#A8E6CF", "#FF8B94", "#95E1D3", "#F38181")
```

<style>
/* Import comic fonts */
@import url('https://fonts.googleapis.com/css2?family=Bangers&family=Comic+Neue:wght@400;700&family=Permanent+Marker&display=swap');

/* Apply comic styling to vignette */
body {
  font-family: 'Comic Neue', 'Comic Sans MS', cursive, sans-serif !important;
  background-color: #FFF9E6;
}

h1, h2, h3, h4 {
  font-family: 'Bangers', cursive !important;
  text-transform: uppercase;
  letter-spacing: 2px;
  color: #E74C3C;
  text-shadow: 2px 2px 0px rgba(0,0,0,0.2);
}

h1 {
  text-align: center;
  padding: 1rem;
  background: linear-gradient(135deg, #FFD93D 0%, #FF6B6B 100%);
  border: 4px solid #000;
  border-radius: 10px;
  box-shadow: 5px 5px 0px rgba(0, 0, 0, 0.3);
  color: #2C3E50 !important;
}

h2 {
  color: #3498DB !important;
  border-bottom: 4px solid #000;
  padding-bottom: 0.5rem;
}

h2::before {
  content: "★ ";
  color: #FFD93D;
}

h3 {
  color: #E67E22 !important;
  border-left: 6px solid #E67E22;
  padding-left: 12px;
}

/* Speech bubble blockquotes */
blockquote {
  background: #FFE66D !important;
  border: 3px solid #000 !important;
  border-radius: 20px !important;
  padding: 1.2rem !important;
  margin: 1.5rem 1rem 1.5rem 2rem !important;
  font-style: italic;
  box-shadow: 4px 4px 0px rgba(0, 0, 0, 0.3);
  position: relative;
}

blockquote::before {
  content: "";
  position: absolute;
  left: -15px;
  top: 15px;
  width: 0;
  height: 0;
  border-right: 15px solid #000;
  border-top: 8px solid transparent;
  border-bottom: 8px solid transparent;
}

blockquote::after {
  content: "";
  position: absolute;
  left: -10px;
  top: 18px;
  width: 0;
  height: 0;
  border-right: 10px solid #FFE66D;
  border-top: 5px solid transparent;
  border-bottom: 5px solid transparent;
}

/* Code blocks - comic border wrapper only, preserve readability */
div.sourceCode {
  border: 3px solid #000 !important;
  border-radius: 8px !important;
  box-shadow: 5px 5px 0px rgba(0, 0, 0, 0.3);
  margin: 1rem 0 !important;
  overflow: hidden;
}

/* Reset pre inside sourceCode to not double-border */
div.sourceCode pre {
  border: none !important;
  box-shadow: none !important;
  margin: 0 !important;
  border-radius: 0 !important;
}

/* Ensure code is readable - don't mess with syntax highlighting */
pre, code {
  font-family: 'SFMono-Regular', Menlo, Monaco, Consolas, monospace !important;
}

/* Inline code - subtle styling */
:not(pre) > code {
  background-color: #f0f0f0 !important;
  padding: 2px 6px !important;
  border-radius: 4px !important;
  border: 1px solid #ddd !important;
}

/* Comic callout boxes */
.callout-kapow {
  background: linear-gradient(135deg, #FF6B6B 0%, #FFD93D 100%);
  border: 4px solid #000;
  border-radius: 15px;
  padding: 1.5rem;
  margin: 1.5rem 0;
  box-shadow: 6px 6px 0px rgba(0, 0, 0, 0.3);
}

.callout-kapow h4 {
  font-family: 'Bangers', cursive !important;
  font-size: 1.8rem;
  text-align: center;
  color: #000 !important;
  text-shadow: 2px 2px 0 #fff;
  margin-bottom: 1rem;
}

.callout-zap {
  background: #A8E6CF;
  border: 4px solid #000;
  border-radius: 15px;
  padding: 1.5rem;
  margin: 1.5rem 0;
  box-shadow: 5px 5px 0px rgba(0, 0, 0, 0.3);
}

.callout-pow {
  background: #95E1D3;
  border: 4px solid #000;
  border-radius: 15px;
  padding: 1.5rem;
  margin: 1.5rem 0;
  box-shadow: 5px 5px 0px rgba(0, 0, 0, 0.3);
}

/* Tables with comic style */
table {
  border: 3px solid #000 !important;
  box-shadow: 4px 4px 0px rgba(0, 0, 0, 0.2);
  margin: 1rem auto;
}

thead {
  background: linear-gradient(135deg, #FF6B6B 0%, #FFD93D 100%) !important;
  font-family: 'Bangers', cursive !important;
}

th, td {
  border: 2px solid #000 !important;
  padding: 10px !important;
}

tbody tr:nth-child(even) {
  background-color: #FFF9E6;
}

tbody tr:hover {
  background-color: #FFE66D;
}

/* Images with comic borders - exclude logos */
img:not([src*="logo"]):not([alt*="logo"]) {
  border: 4px solid #000;
  box-shadow: 5px 5px 0px rgba(0, 0, 0, 0.3);
  background: white;
  padding: 8px;
  border-radius: 5px;
}

/* Ensure logo is not styled */
img[src*="logo"], img[alt*="logo"] {
  border: none !important;
  box-shadow: none !important;
  padding: 0 !important;
  background: transparent !important;
}

/* Strong and emphasis */
strong {
  color: #E74C3C;
}

em {
  color: #3498DB;
}

/* Horizontal rule */
hr {
  border: none;
  height: 4px;
  background: repeating-linear-gradient(
    90deg,
    #000 0px,
    #000 10px,
    transparent 10px,
    transparent 20px
  );
  margin: 2rem 0;
}

/* Comic emphasis spans */
.kapow {
  display: inline-block;
  background: #FFD93D;
  border: 3px solid #000;
  padding: 3px 12px;
  transform: rotate(-2deg);
  font-family: 'Bangers', cursive;
  font-size: 1.1rem;
  box-shadow: 2px 2px 0px rgba(0, 0, 0, 0.3);
}

.zap {
  display: inline-block;
  background: #4ECDC4;
  border: 3px solid #000;
  padding: 3px 12px;
  transform: rotate(1deg);
  font-family: 'Bangers', cursive;
  font-size: 1.1rem;
  box-shadow: 2px 2px 0px rgba(0, 0, 0, 0.3);
}

/* List styling */
ul li::marker {
  color: #E74C3C;
  font-size: 1.2em;
}
</style>

> "The only permanent rule is that you can't play it the same way twice!" - Calvin

Welcome to the first comprehensive analysis of the **Calvinball League**, where chaos meets competition and rules are made up on the spot!

<div class="callout-kapow">
<h4>⚡ KAPOW! ⚡</h4>

Get ready for an exploratory data analysis that examines:

- 🎭 Player performance metrics (including the all-important *style points*!)
- 🏆 Team dynamics and competitive balance
- 🎲 The beautifully unpredictable nature of the game
- 📊 Statistical patterns in pure chaos

*Warning: Traditional sports analytics may not apply here!*
</div>

## Loading the Data

The calvinball package provides four datasets that can be loaded directly:

```{r load-packages}
library(calvinball)
library(dplyr)
library(tidyr)
library(ggplot2)

# Set a comic-inspired theme
theme_set(theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#E74C3C"),
    plot.subtitle = element_text(color = "#666", face = "italic"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(fill = NA, color = "#000", linewidth = 1.5),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom"
  ))
```

```{r load-data}
# The datasets are loaded automatically with the package
# cb_players, cb_teams, cb_games, cb_player_stats

# Quick overview
cat(sprintf("📊 Dataset Overview

• %d players across %d teams
• %d games played over %d seasons
• %d individual performance records

The Calvinball League data is ready for analysis!",
  nrow(cb_players),
  nrow(cb_teams),
  nrow(cb_games),
  length(unique(cb_games$season)),
  nrow(cb_player_stats)
))
```

## League Structure

### Teams

```{r teams-overview}
teams_with_size <- cb_players |>
  count(team_id, name = "roster_size") |>
  left_join(cb_teams, by = "team_id")

teams_with_size |>
  select(team_name, roster_size) |>
  knitr::kable(col.names = c("Team Name", "Roster Size"))
```

### Player Distribution

```{r player-distribution, fig.height=4}
cb_players |>
  left_join(cb_teams, by = "team_id") |>
  count(team_name) |>
  ggplot(aes(x = reorder(team_name, n), y = n, fill = team_name)) +
  geom_col(show.legend = FALSE, color = "#000", linewidth = 1) +
  geom_text(aes(label = n), hjust = -0.3, fontface = "bold") +
  coord_flip() +
  labs(title = "Team Roster Sizes",
       subtitle = "Players are distributed across all teams",
       x = NULL,
       y = "Number of Players") +
  scale_fill_manual(values = comic_colors) +
  ylim(0, max(table(cb_players$team_id)) + 1)
```

## Game Analysis

### Scoring Chaos

<div class="callout-zap">
**⚡ ZAP!** One of the hallmarks of Calvinball is its completely unpredictable scoring system! Scores can range from deeply negative to astronomical!
</div>

```{r score-distribution}
games_long <- cb_games |>
  select(game_id, season, score_home, score_away) |>
  pivot_longer(cols = c(score_home, score_away),
               names_to = "location",
               values_to = "score")

ggplot(games_long, aes(x = score)) +
  geom_histogram(bins = 50, fill = "#3498DB", alpha = 0.8, color = "#000") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#E74C3C", linewidth = 1.5) +
  labs(title = "Distribution of Game Scores",
       subtitle = "Note the negative scores and extreme outliers - pure Calvinball!",
       x = "Score",
       y = "Frequency")
```

```{r score-summary}
score_stats <- games_long |>
  summarise(
    Minimum = min(score),
    Median = median(score),
    Mean = round(mean(score), 1),
    Maximum = max(score),
    `Std Dev` = round(sd(score), 1)
  )

knitr::kable(score_stats, caption = "📊 Score Summary Statistics")
```

### Scoring Types

In true Calvinball fashion, the scoring system changes from game to game!

```{r scoring-types}
cb_games |>
  count(scoring_type) |>
  ggplot(aes(x = reorder(scoring_type, n), y = n, fill = scoring_type)) +
  geom_col(show.legend = FALSE, color = "#000", linewidth = 1) +
  geom_text(aes(label = n), hjust = -0.3, size = 3.5, fontface = "bold") +
  coord_flip() +
  labs(title = "Frequency of Scoring Types",
       subtitle = "Because consistency is boring!",
       x = NULL,
       y = "Number of Games") +
  scale_fill_manual(values = comic_colors) +
  ylim(0, max(table(cb_games$scoring_type)) + 3)
```

## Team Performance

### Win-Loss Records

First, let's compute team records from the games data:

```{r compute-team-records}
# Compute team records from games
team_records <- cb_games |>
  select(season, team_id = team_home, score_for = score_home, score_against = score_away) |>
  bind_rows(
    cb_games |>
      select(season, team_id = team_away, score_for = score_away, score_against = score_home)
  ) |>
  mutate(
    win = score_for > score_against,
    loss = score_for < score_against,
    tie = score_for == score_against
  ) |>
  group_by(team_id, season) |>
  summarize(
    wins = sum(win),
    losses = sum(loss),
    ties = sum(tie),
    .groups = "drop"
  ) |>
  left_join(cb_teams, by = "team_id")
```

```{r team-records, fig.height=6}
team_records_plot <- team_records |>
  group_by(team_name) |>
  mutate(total_wins = sum(wins)) |>
  ungroup() |>
  mutate(season = factor(season))

ggplot(team_records_plot, aes(x = reorder(team_name, total_wins), y = wins, fill = season)) +
  geom_col(position = "dodge", color = "#000", linewidth = 0.5) +
  coord_flip() +
  labs(title = "Team Wins by Season",
       subtitle = "How do teams stack up across seasons?",
       x = NULL,
       y = "Wins",
       fill = "Season") +
  scale_fill_manual(values = c("#FF6B6B", "#4ECDC4", "#FFD93D"))
```

```{r overall-records}
overall_records <- team_records |>
  group_by(team_id, team_name) |>
  summarise(
    total_wins = sum(wins),
    total_losses = sum(losses),
    total_ties = sum(ties),
    .groups = "drop"
  ) |>
  mutate(
    total_games = total_wins + total_losses + total_ties,
    win_pct = round(total_wins / total_games, 3)
  ) |>
  arrange(desc(win_pct))

overall_records |>
  select(team_name, total_wins, total_losses, total_ties, win_pct) |>
  knitr::kable(col.names = c("Team", "Wins", "Losses", "Ties", "Win %"),
               caption = "🏆 Overall Team Records (All Seasons)")
```

### Home vs Away Performance

```{r home-away}
home_away_summary <- cb_games |>
  summarise(
    `Home Wins` = sum(score_home > score_away),
    `Away Wins` = sum(score_away > score_home),
    `Ties` = sum(score_home == score_away)
  ) |>
  pivot_longer(everything(), names_to = "outcome", values_to = "count")

ggplot(home_away_summary, aes(x = outcome, y = count, fill = outcome)) +
  geom_col(show.legend = FALSE, color = "#000", linewidth = 1) +
  geom_text(aes(label = count), vjust = -0.5, size = 6, fontface = "bold") +
  labs(title = "Home vs Away Performance",
       subtitle = "Surprisingly, there might be a home field advantage!",
       x = NULL,
       y = "Number of Games") +
  scale_fill_manual(values = c("Home Wins" = "#27AE60",
                                "Away Wins" = "#E74C3C",
                                "Ties" = "#95A5A6")) +
  ylim(0, max(home_away_summary$count) + 10)
```

## Player Performance

### Computing Player Summaries

Let's summarize player statistics from the individual game data:

```{r compute-player-summary}
player_summary <- cb_player_stats |>
  group_by(player_id) |>
  summarize(
    games_played = n(),
    avg_wickets = mean(wickets_scored),
    avg_opposite_touchdowns = mean(opposite_touchdowns),
    avg_style_points = mean(style_points),
    total_rule_declarations = sum(spontaneous_rule_declarations),
    avg_time_played = mean(minutes_played),
    .groups = "drop"
  ) |>
  left_join(cb_players, by = "player_id")
```

### Style Points Leaders

<div class="callout-pow">
**⭐ STYLE MATTERS!** In Calvinball, style is just as important as substance. These are the players with the most panache!
</div>

```{r style-leaders}
top_style <- player_summary |>
  arrange(desc(avg_style_points)) |>
  head(10) |>
  left_join(cb_teams, by = "team_id")

top_style |>
  select(player_name, team_name, games_played, avg_style_points) |>
  mutate(avg_style_points = round(avg_style_points, 2)) |>
  knitr::kable(col.names = c("Player", "Team", "Games", "Avg Style Points"),
               caption = "⭐ Top 10 Players by Average Style Points")
```

```{r style-plot, fig.height=6}
top_style |>
  ggplot(aes(x = reorder(player_name, avg_style_points),
             y = avg_style_points,
             fill = team_name)) +
  geom_col(color = "#000", linewidth = 1) +
  coord_flip() +
  labs(title = "Top 10 Players by Average Style Points",
       subtitle = "Flair and panache are everything in Calvinball!",
       x = NULL,
       y = "Average Style Points",
       fill = "Team") +
  scale_fill_manual(values = comic_colors)
```

### Wickets and Performance Metrics

```{r wickets-analysis}
player_summary |>
  ggplot(aes(x = avg_wickets, y = avg_style_points)) +
  geom_point(alpha = 0.7, size = 4, color = "#3498DB") +
  geom_smooth(method = "lm", se = TRUE, color = "#E74C3C", linetype = "dashed", linewidth = 1.5) +
  labs(title = "Style Points vs Wickets Scored",
       subtitle = "Is there a relationship between technical skill and style?",
       x = "Average Wickets Scored",
       y = "Average Style Points")
```

### Rule Declarations

<div class="callout-zap">
**🎨 CREATIVE GENIUS!** Spontaneous rule declarations are a cornerstone of Calvinball strategy! The most innovative players make up the best rules!
</div>

```{r rule-declarations}
top_declarers <- player_summary |>
  arrange(desc(total_rule_declarations)) |>
  head(10) |>
  left_join(cb_teams, by = "team_id")

top_declarers |>
  select(player_name, team_name, total_rule_declarations, games_played) |>
  knitr::kable(col.names = c("Player", "Team", "Total Declarations", "Games Played"),
               caption = "🎨 Top Rule Declarers - The Most Creative Minds!")
```

## Advanced Metrics

### Player Versatility Score

Let's create a versatility metric combining multiple stats:

```{r versatility, fig.height=7}
player_versatility <- player_summary |>
  left_join(cb_teams, by = "team_id") |>
  mutate(
    # Normalize each metric to 0-1 scale
    style_norm = (avg_style_points - min(avg_style_points)) /
                 (max(avg_style_points) - min(avg_style_points)),
    wickets_norm = (avg_wickets - min(avg_wickets)) /
                   (max(avg_wickets) - min(avg_wickets)),
    declarations_norm = (total_rule_declarations - min(total_rule_declarations)) /
                        (max(total_rule_declarations) - min(total_rule_declarations)),
    # Versatility score
    versatility = (style_norm + wickets_norm + declarations_norm) / 3
  ) |>
  arrange(desc(versatility)) |>
  head(15)

player_versatility |>
  ggplot(aes(x = reorder(player_name, versatility),
             y = versatility,
             fill = team_name)) +
  geom_col(color = "#000", linewidth = 1) +
  coord_flip() +
  labs(title = "Top 15 Most Versatile Players",
       subtitle = "Combining style, wickets, and rule declarations",
       x = NULL,
       y = "Versatility Score",
       fill = "Team") +
  scale_fill_manual(values = comic_colors)
```

## Generating Fresh Data

If you want to generate your own Calvinball data with different parameters, use the `generate_calvinball_data()` function:

```{r generate-example, eval=FALSE}
# Generate a smaller league
small_league <- generate_calvinball_data(
  n_players = 12,
  n_teams = 3,
  n_games = 20,
  n_seasons = 1
)

# Access the datasets
small_league$players
small_league$games
```

---

<div class="callout-kapow">
<h4>💥 BAM! Key Findings 💥</h4>

**1. SCORING CHAOS!** Calvinball lives up to its reputation with scores ranging from `r min(games_long$score)` to `r format(max(games_long$score), big.mark = ",")`, with a standard deviation of `r format(round(sd(games_long$score), 1), big.mark = ",")`. That's WILD!

**2. HOME FIELD ADVANTAGE?!** Despite the chaotic nature of the game, there appears to be a slight home field advantage, with home teams winning approximately `r round(100 * mean(cb_games$score_home > cb_games$score_away), 1)`% of games. Who knew?

**3. STYLE MATTERS!** The top performers in style points show considerable variability, suggesting that multiple approaches to Calvinball can be successful.

**4. COMPETITIVE BALANCE!** The league shows good parity, with win percentages ranging from `r round(min(overall_records$win_pct), 3)` to `r round(max(overall_records$win_pct), 3)`. Anyone can win!

**5. CREATIVE GENIUS!** Players average `r round(mean(player_summary$total_rule_declarations, na.rm = TRUE), 1)` spontaneous rule declarations over their careers, demonstrating the inventive spirit of the game.
</div>

---

<center>
*"Sometimes I think the surest sign that intelligent life exists elsewhere in the universe is that none of it has tried to contact us."*

<span class="kapow">🏴 The Score is STILL Q to 12! 🏴</span>
</center>
